---
title: "Beyond Zipf's Law: a reproducible generative benchmark"
subtitle: "v0.5.1 — stable latent calibration, Markov control, MLE robustness, bootstrap AUC, joint matching, and NMI sensitivity"
author: "Working reproducibility notebook"
date: last-modified
bibliography: references.bib
params:
  run_simulations: true
  force_rerun: false
  quick: false
  n_rep: 100
  n_cal_seeds: 10
  n_latent_cal_seeds: 50
  n_surrogates: 20
  n_boot: 2000
  heaps_window: 7
  nmi_top_k_values: "25,50,100"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
jupyter: false
---

## Purpose of v0.5.1

Version 0.5.1 freezes the benchmark architecture introduced in v0.5 and makes one prespecified numerical correction: the **latent-mixture parameter is selected on a 50-seed calibration panel** rather than on 10 seeds. The held-out evaluation panel remains seeds 2001 onward and is never used for calibration.

The notebook addresses five robustness questions raised during review:

1. **Sequential control:** a stationary non-SSR Markov process has the same finite-Zipf marginal as the exchangeable reference but non-zero temporal dependence.
2. **Estimator robustness:** the deliberately simple OLS log--log exponent is compared with a conditional discrete-likelihood estimate on the identical rank window.
3. **Uncertainty in discrimination:** pairwise AUC effects are accompanied by bootstrap intervals for the signed rank effect $R=2\operatorname{AUC}-1$.
4. **Joint matching:** a second benchmark matches $(\widehat\alpha,V_n)$ before evaluating diagnostics not used in calibration.
5. **NMI sensitivity:** excess lag-1 normalized mutual information is recomputed after retaining $K=25,50,100$ frequent types plus one `OTHER` category.

The central inferential hierarchy is therefore

$$
\text{marginal scaling}
\longrightarrow
\text{order dependence}
\longrightarrow
\text{directional/history-dependent structure}.
$$

```{r}
#| label: setup
required <- c("ggplot2", "patchwork", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Install required packages first: ", paste(missing, collapse = ", "))
}

source("R/generators.R")
source("R/diagnostics.R")
source("R/calibration.R")
source("R/figures.R")
source("R/run_all.R")

dir.create("figures", showWarnings = FALSE, recursive = TRUE)
dir.create("results", showWarnings = FALSE, recursive = TRUE)
dir.create("tables", showWarnings = FALSE, recursive = TRUE)
set.seed(20260808)

parse_integer_vector <- function(x) {
  as.integer(trimws(strsplit(as.character(x), ",", fixed = TRUE)[[1L]]))
}
nmi_top_k_values <- parse_integer_vector(params$nmi_top_k_values)

fmt_mean_sd <- function(mean, sd, digits = 3L) {
  sprintf(paste0("%.", digits, "f (%.", digits, "f)"), mean, sd)
}

save_table <- function(df, stem, caption = NULL, digits = 3L) {
  csv_path <- file.path("tables", paste0(stem, ".csv"))
  tex_path <- file.path("tables", paste0(stem, ".tex"))
  utils::write.csv(df, csv_path, row.names = FALSE)
  latex <- knitr::kable(
    df, format = "latex", booktabs = TRUE, digits = digits,
    caption = caption, escape = TRUE
  )
  writeLines(as.character(latex), tex_path)
  knitr::kable(df, format = "html", digits = digits, caption = caption)
}
```

## One-click use in RStudio

A fresh copy of this repository can be rendered directly in RStudio. With the default settings above, the first **Render** runs the full paper analysis because the required result files do not yet exist. Later renders detect those files and rebuild only the notebook, tables and figures.

To deliberately rerun every calibration and held-out replication, set

```yaml
force_rerun: true
```

and render once. Set it back to `false` afterwards.

```{r}
#| label: optional-run
should_run <- isTRUE(params$run_simulations) &&
  (isTRUE(params$force_rerun) || !paper_outputs_exist("results"))

if (should_run) {
  if (isTRUE(params$quick)) {
    run_all(
      n_rep = 10L,
      n_cal_seeds = 3L,
      n_latent_cal_seeds = 8L,
      n_surrogates = 3L,
      heaps_window = as.integer(params$heaps_window),
      n_boot = 250L,
      nmi_top_k_values = nmi_top_k_values
    )
  } else {
    run_all(
      n_rep = as.integer(params$n_rep),
      n_cal_seeds = as.integer(params$n_cal_seeds),
      n_latent_cal_seeds = as.integer(params$n_latent_cal_seeds),
      n_surrogates = as.integer(params$n_surrogates),
      heaps_window = as.integer(params$heaps_window),
      n_boot = as.integer(params$n_boot),
      nmi_top_k_values = nmi_top_k_values
    )
  }
} else if (paper_outputs_exist("results")) {
  cat("Existing paper outputs found: simulations were not rerun.\n")
}
```

```{r}
#| label: require-results
needed_results <- .paper_outputs("results")
missing_results <- needed_results[!file.exists(needed_results)]
if (length(missing_results)) {
  stop(
    "Missing simulation outputs: ", paste(missing_results, collapse = ", "),
    ". Set run_simulations: true (or force_rerun: true) and Render once."
  )
}
```

# 1. Calibration design

## 1.1 Stable latent-mixture calibration

The latent construction showed greater Monte Carlo variability than the other benchmark constructions. Version 0.5.1 therefore uses a two-stage procedure that remains completely separate from held-out evaluation:

- the broad latent grid is screened on a small prespecified subset of calibration seeds;
- the best regions and local refinements are evaluated on **50 calibration seeds**;
- the final latent parameter is selected from this 50-seed calibration panel;
- evaluation still uses seeds 2001--2100.

This change increases calibration precision without using any held-out information for tuning.

```{r}
#| label: calibration-tables
cal_exp <- readRDS("results/calibration_exponent.rds")
cal_joint <- readRDS("results/calibration_joint.rds")
cal_exp_summary <- utils::read.csv("results/calibration_exponent_summary.csv", stringsAsFactors = FALSE)
cal_joint_summary <- utils::read.csv("results/calibration_joint_summary.csv", stringsAsFactors = FALSE)

save_table(
  cal_exp_summary,
  "table_calibration_exponent_v051",
  caption = "Exponent-only calibration. The latent-mixture final parameter is selected on the enlarged calibration panel."
)
```

```{r}
#| label: calibration-joint-table
save_table(
  cal_joint_summary,
  "table_calibration_joint_v051",
  caption = "Joint calibration targeting the OLS exponent and final vocabulary size."
)
```

# 2. Generative constructions and observation map

The strict benchmark contains four constructions with deliberately different dependence structures.

**Finite Zipf reference.** Independent tokens are sampled from

$$
p(r;1,V)=\frac{r^{-1}}{\sum_{j=1}^V j^{-1}}.
$$

It is a transparent constraint-based reference rather than a dynamical mechanism.

**Zipf Markov control.** The stationary marginal is the same finite-Zipf distribution, while transitions follow

$$
P_{ij}=\rho\mathbf 1\{i=j\}+(1-\rho)p_j,
\qquad \rho=0.35.
$$

It therefore provides a non-SSR process with genuine serial dependence.

**Latent mixture.** Sequences are conditionally non-Zipfian but heterogeneous in scale; pooling can create an approximately Zipfian marginal [@schwab2014; @aitchison2016].

**Sample-space reduction.** The current state restricts the accessible future state space until restart [@corominas2015; @thurner2015].

Random segmentation and Simon reinforcement are retained separately as observation-window / compatibility examples [@simon1955; @li1992; @conrad2004; @ferrerElvevag2010].

The prespecified primary rank summary remains

$$
10\le r\le500,\qquad N_{(r)}\ge5,
$$

with

$$
\log \widehat p_{(r)}=a-\widehat\alpha_{\mathrm{OLS}}\log r+\varepsilon_r.
$$

This OLS slope is intentionally a low-dimensional descriptive statistic, not a stand-alone power-law goodness-of-fit test [@clauset2009].

The likelihood robustness estimator uses the same retained ranks $\mathcal R$:

$$
q_r(\alpha)=\frac{r^{-\alpha}}{\sum_{j\in\mathcal R}j^{-\alpha}},
$$

$$
\ell(\alpha)
=-\alpha\sum_{r\in\mathcal R}N_{(r)}\log r
-N_{\mathcal R}\log\left(\sum_{j\in\mathcal R}j^{-\alpha}\right).
$$

# 3. Held-out benchmark

```{r}
#| label: load-heldout
rep_strict <- readRDS("results/benchmark_strict_replications.rds")
rep_mimic <- readRDS("results/benchmark_mimicking_replications.rds")
rep_joint <- readRDS("results/benchmark_joint_strict_replications.rds")
auc_boot <- utils::read.csv("results/diagnostic_pairwise_auc_bootstrap_strict.csv", stringsAsFactors = FALSE)
auc_boot_joint <- utils::read.csv("results/diagnostic_pairwise_auc_bootstrap_joint.csv", stringsAsFactors = FALSE)
nmi_topk <- utils::read.csv("results/nmi_topk_sensitivity_strict.csv", stringsAsFactors = FALSE)

strict_parameters <- strict_benchmark_parameters(
  exponent_parameters = cal_exp$selected,
  n_tokens = 100000L,
  maxent_vocabulary_size = 5000L,
  markov_vocabulary_size = 5000L,
  markov_persistence = 0.35
)
sims_strict <- simulate_strict_benchmark_once(
  seed = 2001L, n_tokens = 100000L, parameters = strict_parameters
)
```

## 3.1 Main numerical table

```{r}
#| label: strict-summary-table
strict_summary <- diagnostic_summary_table(rep_strict)
get_diag <- function(g, v, field = "mean") {
  x <- strict_summary[strict_summary$generator == g & strict_summary$diagnostic == v, field]
  if (!length(x)) NA_real_ else x[1L]
}
get_diag_sd <- function(g, v) get_diag(g, v, "sd")

generators <- unique(rep_strict$generator)
strict_wide <- data.frame(
  Construction = generators,
  `OLS alpha` = vapply(generators, function(g) fmt_mean_sd(get_diag(g, "alpha"), get_diag_sd(g, "alpha")), character(1)),
  `MLE alpha` = vapply(generators, function(g) fmt_mean_sd(get_diag(g, "alpha_mle"), get_diag_sd(g, "alpha_mle")), character(1)),
  `Final V` = vapply(generators, function(g) fmt_mean_sd(get_diag(g, "vocabulary"), get_diag_sd(g, "vocabulary"), digits = 1L), character(1)),
  `Excess NMI` = vapply(generators, function(g) fmt_mean_sd(get_diag(g, "transition_nmi_excess"), get_diag_sd(g, "transition_nmi_excess")), character(1)),
  `Direction D` = vapply(generators, function(g) fmt_mean_sd(get_diag(g, "transition_toward_more_frequent"), get_diag_sd(g, "transition_toward_more_frequent")), character(1)),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
save_table(
  strict_wide,
  "table_heldout_strict_v051",
  caption = "Held-out diagnostics (mean and standard deviation across independent replications)."
)
```

## 3.2 Marginal scaling and OLS/MLE robustness

```{r}
#| label: fig-rank-frequency-v051
#| fig-cap: "Distinct constructions can display similar marginal rank--frequency scaling in a held-out realization."
#| fig-width: 7.5
#| fig-height: 5
p_rank <- plot_rank_frequency(sims_strict)
p_rank
ggplot2::ggsave("figures/fig_rank_frequency_v051.pdf", p_rank, width = 7.5, height = 5)
```

```{r}
#| label: fig-alpha-estimators-v051
#| fig-cap: "Held-out rank-exponent estimates under the prespecified OLS summary and conditional discrete likelihood."
#| fig-width: 8
#| fig-height: 5.5
p_estimators <- plot_alpha_estimator_comparison(rep_strict)
p_estimators
ggplot2::ggsave("figures/fig_alpha_estimator_robustness_v051.pdf", p_estimators, width = 8, height = 5.5)
```

# 4. Sequential control: generic order dependence versus SSR directionality

Excess lag-1 normalized mutual information is calculated after retaining the $K$ most frequent types plus `OTHER` and subtracting the expectation under within-sequence shuffling:

$$
I_{\mathrm{excess}}(K)
=I_{\mathrm{NMI}}^{(K)}(X_t;X_{t+1})
-E_{\mathrm{shuffle}}\left[I_{\mathrm{NMI}}^{(K)}(X_t;X_{t+1})\right].
$$

This statistic tests for order dependence. It is deliberately **not** interpreted as SSR-specific. Directionality is measured separately by

$$
D=P(R_{t+1}<R_t\mid R_{t+1}\ne R_t).
$$

For exchangeable pairs, and for the stationary refresh component of the Markov control conditional on a change, symmetry gives $D\simeq1/2$. Canonical SSR instead predicts a strong directional asymmetry.

```{r}
#| label: fig-sequence-tree-v051
#| fig-cap: "The Markov control separates generic sequential dependence from SSR-like directionality: excess information detects order dependence, while transition direction isolates the asymmetric sample-space-reduction signature."
#| fig-width: 12
#| fig-height: 5.5
p_sequence <- plot_sequence_diagnostic_figure(rep_strict)
p_sequence
ggplot2::ggsave("figures/fig_sequence_identification_v051.pdf", p_sequence, width = 12, height = 5.5)
```

# 5. Bootstrap uncertainty in pairwise discrimination

For a scalar diagnostic $T$, pairwise discrimination is summarized by the signed rank effect

$$
R_T=2\operatorname{AUC}_T-1\in[-1,1].
$$

The absolute value $|R_T|$ is a bounded effect size: 0 denotes no ordinal discrimination and 1 perfect separation. We bootstrap the two held-out samples independently and report 95% intervals for the **signed** effect, so near-zero absolute effects are not mistaken for precise evidence of separation.

```{r}
#| label: auc-bootstrap-table
auc_key <- auc_boot[auc_boot$diagnostic %in% c(
  "alpha", "alpha_mle", "transition_nmi_excess", "transition_toward_more_frequent"
), c(
  "pair", "diagnostic", "rank_effect", "rank_effect_lower", "rank_effect_upper",
  "separation", "ci_contains_zero", "n_x", "n_y"
)]
auc_key$rank_effect <- round(auc_key$rank_effect, 3)
auc_key$rank_effect_lower <- round(auc_key$rank_effect_lower, 3)
auc_key$rank_effect_upper <- round(auc_key$rank_effect_upper, 3)
auc_key$separation <- round(auc_key$separation, 3)
save_table(
  auc_key,
  "table_auc_bootstrap_key_v051",
  caption = "Pairwise signed AUC rank effects and 95% bootstrap intervals for the principal marginal and sequential diagnostics."
)
```

```{r}
#| label: fig-bootstrap-key-v051
#| fig-cap: "Bootstrap intervals distinguish weak exponent-based discrimination from strong structural separation."
#| fig-width: 11
#| fig-height: 8
p_boot_key <- plot_bootstrap_selected_diagnostics(auc_boot)
p_boot_key
ggplot2::ggsave("figures/fig_bootstrap_key_diagnostics_v051.pdf", p_boot_key, width = 11, height = 8)
```

# 6. Sensitivity of excess NMI to alphabet collapse

```{r}
#| label: nmi-sensitivity-table
nmi_key <- aggregate(
  transition_nmi_excess ~ generator + top_k,
  data = nmi_topk,
  FUN = mean
)
nmi_wide <- reshape(
  nmi_key,
  idvar = "generator",
  timevar = "top_k",
  direction = "wide"
)
names(nmi_wide) <- sub("transition_nmi_excess\\.", "K=", names(nmi_wide))
for (j in seq_along(nmi_wide)) {
  if (is.numeric(nmi_wide[[j]])) nmi_wide[[j]] <- round(nmi_wide[[j]], 4)
}
save_table(
  nmi_wide,
  "table_nmi_sensitivity_v051",
  caption = "Mean excess lag-1 NMI after retaining the K most frequent types plus OTHER."
)
```

```{r}
#| label: fig-nmi-sensitivity-v051
#| fig-cap: "Sequence-information conclusions are stable when the collapsed alphabet retains 25, 50, or 100 frequent types."
#| fig-width: 8
#| fig-height: 5
p_nmi_k <- plot_nmi_sensitivity(nmi_topk)
p_nmi_k
ggplot2::ggsave("figures/fig_nmi_topk_sensitivity_v051.pdf", p_nmi_k, width = 8, height = 5)
```

# 7. Joint matching on $(\widehat\alpha,V_n)$

The robustness benchmark explicitly targets

$$
(\widehat\alpha_{\mathrm{OLS}},V_n)\approx(1,5000)
$$

using only these two marginal summaries. The relevant question is whether diagnostics **not used in calibration** continue to distinguish the constructions.

```{r}
#| label: joint-summary-table
joint_summary <- diagnostic_summary_table(rep_joint)
get_joint <- function(g, v, field = "mean") {
  x <- joint_summary[joint_summary$generator == g & joint_summary$diagnostic == v, field]
  if (!length(x)) NA_real_ else x[1L]
}
joint_gens <- unique(rep_joint$generator)
joint_wide <- data.frame(
  Construction = joint_gens,
  `OLS alpha` = vapply(joint_gens, function(g) fmt_mean_sd(get_joint(g, "alpha"), get_joint(g, "alpha", "sd")), character(1)),
  `Final V` = vapply(joint_gens, function(g) fmt_mean_sd(get_joint(g, "vocabulary"), get_joint(g, "vocabulary", "sd"), digits = 1L), character(1)),
  `Excess NMI` = vapply(joint_gens, function(g) fmt_mean_sd(get_joint(g, "transition_nmi_excess"), get_joint(g, "transition_nmi_excess", "sd")), character(1)),
  `Direction D` = vapply(joint_gens, function(g) fmt_mean_sd(get_joint(g, "transition_toward_more_frequent"), get_joint(g, "transition_toward_more_frequent", "sd")), character(1)),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
save_table(
  joint_wide,
  "table_joint_matching_v051",
  caption = "Held-out diagnostics after joint calibration on the OLS exponent and final vocabulary size."
)
```

```{r}
#| label: fig-joint-robustness-v051
#| fig-cap: "Jointly matching the marginal exponent and final vocabulary removes a trivial marginal discriminator while leaving order-sensitive structure visible."
#| fig-width: 12
#| fig-height: 5.5
p_joint_robust <- plot_joint_sequence_robustness(rep_joint)
p_joint_robust
ggplot2::ggsave("figures/fig_joint_matching_sequence_v051.pdf", p_joint_robust, width = 12, height = 5.5)
```

# 8. Observation-window dependence and Zipf mimicry

The compatible/mimicking benchmark is not forced into the strict equivalence class. Instead it tests whether the fitted exponent itself changes materially with the analyst's rank window.

```{r}
#| label: observation-window-data
strict_sens <- utils::read.csv("results/alpha_sensitivity_strict_replications.csv", stringsAsFactors = FALSE)
mimic_sens <- utils::read.csv("results/alpha_sensitivity_mimicking_replications.csv", stringsAsFactors = FALSE)
alpha_sens_all <- rbind(strict_sens, mimic_sens)
alpha_window_summary <- summarise_alpha_sensitivity_replications(alpha_sens_all)
utils::write.csv(
  alpha_window_summary,
  "tables/table_alpha_window_summary_v051.csv",
  row.names = FALSE
)
```

```{r}
#| label: fig-window-v051
#| fig-cap: "The fitted Zipf exponent can depend strongly on the requested rank window; open points denote windows truncated by the minimum-count rule."
#| fig-width: 9
#| fig-height: 5.5
p_window <- plot_alpha_sensitivity(alpha_sens_all, primary_r_max = 500L)
p_window
ggplot2::ggsave("figures/fig_alpha_window_sensitivity_v051.pdf", p_window, width = 9, height = 5.5)
```

# 9. Latent conditioning

The latent mixture is not identified by temporal diagnostics because it is exchangeable within sequence. Its informative diagnostic is instead **conditioning**: the pooled marginal may scale while scale-conditioned rank distributions do not.

```{r}
#| label: fig-latent-conditioning-v051
#| fig-cap: "Pooling across latent scales can create marginal scaling absent within scale-conditioned regimes."
#| fig-width: 7.5
#| fig-height: 5
p_latent <- plot_latent_conditioning(sims_strict$latent)
p_latent
ggplot2::ggsave("figures/fig_latent_conditioning_v051.pdf", p_latent, width = 7.5, height = 5)
```

# 10. Outputs for the paper

The main paper-facing outputs produced by this notebook are:

```text
TABLES
  tables/table_calibration_exponent_v051.{csv,tex}
  tables/table_calibration_joint_v051.{csv,tex}
  tables/table_heldout_strict_v051.{csv,tex}
  tables/table_auc_bootstrap_key_v051.{csv,tex}
  tables/table_nmi_sensitivity_v051.{csv,tex}
  tables/table_joint_matching_v051.{csv,tex}

FIGURES
  figures/fig_rank_frequency_v051.pdf
  figures/fig_alpha_estimator_robustness_v051.pdf
  figures/fig_sequence_identification_v051.pdf
  figures/fig_bootstrap_key_diagnostics_v051.pdf
  figures/fig_nmi_topk_sensitivity_v051.pdf
  figures/fig_joint_matching_sequence_v051.pdf
  figures/fig_alpha_window_sensitivity_v051.pdf
  figures/fig_latent_conditioning_v051.pdf
```

The benchmark supports a deliberately limited inferential claim: a fitted Zipf exponent can leave very different constructions observationally difficult to distinguish under one marginal summary, while conditioning, vocabulary accumulation, and temporal diagnostics can progressively separate them. The Markov control is essential because it prevents generic serial dependence from being mistaken for an SSR-specific signature.

```{r}
#| label: session-info
session_text <- capture.output(sessionInfo())
writeLines(session_text, "results/sessionInfo.txt")
sessionInfo()
```
